Future challenges for Research Software Engineering

### 2025-09-18 correspondence/rebuttal-1.md

> 3. The authors claim to "present here the results of a workshop discussion at the deRSE25 conference in Karlsruhe". However, they also include their hopes, discussions and own expectations for the future RSE ecosystem (in Germany). This should be mentioned in the introduction and it should be made clear which statements have been put forward during the discussion and which have been added by the authors after the discussion. The authors mention this construct in section 3 ("In the previous sections, we have tried to list some of the trends that we foresee. The material presented here is based on the outputs of a workshop, but extended with the authors’ ideas."), which makes understanding the paper difficult; I would expect such a statement at the beginning. While I do not propose to re-structure the paper (e.g. report first, discussion second), I suggest to review the paper as to which statements have been reported in the discussion and which statement have been brought up by the authors.

We have put this sentence to the front, and kept it at the end as a reminder.

We have reworded the text as suggested.

We have moved this paragraph further down.

> "What issues have newly emerged, where they are now just happy to say: .." This seems

Not clear what was meant here. We have reworded it nevertheless.

> 2.1 Chapter
> "might remain unchanged, there might be a shift in the responsibilities of RSEs, in the form of a partial reallocation of the training budget to the other core competencies. " What is the rationale for this statement?

We reprased it to better reflect that adaptations will be necessary

Clarified that we mean the competencies.

> Chapter 2.2
> "visibility and accelerate the professionalisation of the RSE role" -> missing line spacing

fixed this.

Reworded it to keep the ambiguity in Gender.

@@

> Chapter 2.2
> "visibility and accelerate the professionalisationof the RSE role" -> missing line spacing

> Story Box 4
> "Kim accelerates his research.." To my understanding, Kim and Kay are female.

> Chapter 2.5
> "As engineers, RSEs take the responsibility for the software they develop. RSEs need to be aware of these issues and develop their own “red lines”." This statement underestimates the complexity of how code/knowledge/algorithms can be reused for military purposes. A scientist would need to know about present military technologies which is unrealistic. I fear the call for responsibility is too much to ask from an individual. Additionally, it creates a huge burden if code or a programme has been reused for military purposes. I would suggest that the authors make clear the boundaries of the responsibility.

Thank you for the comment. We have changed the sentence to point to the dual use dilemma.

@@

> Story Box 6:
> "that are backed by large, foreign corporate entities. So far, she still has the support of the university board, since it values strategic autonomy." This citation shows that the authors seem to write from a German point of view. In the previous sections, international collaboration was highlighted. In a global world, there is no "foreign corporate entity", just collaboration tools from different sources. This is a minor comment here but important for the general remark regarding the perspective given in the paper.

Good point, we have dropped foreign as it does not matter where a large cooperation has its head quarters

> Chapter 2.7
> "in lower tiers of an of a research organisation", sentence needs attention

### 2025-09-18 future_challenges_for_RSEng.tex
resolve box 2

While the RSE core competencies~\cite{Goth2024} might remain unchanged,
there might be a shift in the responsibilities of RSEs,
in the form of a partial reallocation of the training budget
to the other core competencies.
Regarding the definition of RSEs, we will still have a spectrum of competencies,
from full professional RSEs to domain scientists who develop research software.
We note a similar discussion took place regarding how SE roles had to adapt to
rapid changes brought by automation, cloud services and new project management
methodologies~\cite{Meade2019}.

@@

Looking at that on a broader scale, especially the education aspect deserves some attention.\\
The Covid-19 pandemic acted as a catalyst for the digital transformation
of higher education~\cite{Bygstad2022}. Students and early-career researchers
are now more familiar with digital learning tools, which lead to a shift
in the responsibilities of teachers. To quote from the original study,
``with so many digital resources at hand,
the task of the lecturer will be fewer lectures,
and to act more as a facilitator of resources,
and to monitor activities and results over time.''~\cite{Bygstad2022}

@@

\begin{story}{The interaction of RSEs with AI}{RSEs_with_AI}
Kim develops advanced scientific software in tandem with an AI agent that codes,
debugs, and simulates experiments alongside him. The AI proactively suggests optimisations,
generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
With this assistance, Kim accelerates his research output while maintaining full control over scientific direction and ethical oversight.
\end{story}

>>>>

While the RSE core competencies~\cite{Goth2024} might remain unchanged,
there might be a shift in the responsibilities of RSEs and thereby requiring a realignment of the priorities in training.
Regarding the definition of RSEs, we will still have a spectrum of competencies,
from full professional RSEs to domain scientists who develop research software.
We note a similar discussion took place regarding how SE roles had to adapt to
rapid changes brought by automation, cloud services and new project management
methodologies~\cite{Meade2019}.

@@

Looking at the competencies on a broader scale, especially the education aspect deserves some attention.\\
The Covid-19 pandemic acted as a catalyst for the digital transformation
of higher education~\cite{Bygstad2022}. Students and early-career researchers
are now more familiar with digital learning tools, which lead to a shift
in the responsibilities of teachers. To quote from the original study,
``with so many digital resources at hand,
the task of the lecturer will be fewer lectures,
and to act more as a facilitator of resources,
and to monitor activities and results over time.''~\cite{Bygstad2022}

@@

\begin{story}{The interaction of RSEs with AI}{RSEs_with_AI}
Kim develops advanced scientific software in tandem with an AI agent that codes,
debugs, and simulates experiments alongside him. The AI proactively suggests optimisations,
generates reproducible workflows, and adapts to Kim’s coding style and domain-specific needs.
With this assistance, Kim's research output is amplified while maintaining full control over the scientific direction and ethical oversight.
\end{story}
